---
title: "Phenology data - visualization and simple models"
author: "Susannah Tysor"
date: "April 16, 2019"
output: html_document
editor_options:
    chunk_output_type: console
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)
```

```{r depends}
library(ggplot2)
library(dplyr)
library(tidyr)
library(summarytools)
library(arm)

```
This document explores the phenology data and forcing in a simple way in order to clarify modelling decisions in stan.

1) Should I use ristos or 5 degree heatsum threshold?
2) Does slope vary by transition state?
3) Do I need to include slope effects in addition to intercept effects?


```{r data, echo=FALSE, results="asis"}


phenology_data <- read.csv("data/stan_input/phenology_heatsum.csv", stringsAsFactors = FALSE, header = TRUE) 

SPU_dat <- read.csv("~/Documents/research_phd/data/OrchardInfo/LodgepoleSPUs.csv", header=TRUE, stringsAsFactors = FALSE) %>%
    dplyr::select(SPU_Name, Orchard) #add provenance information

phendf <- phenology_data %>%
    na.omit()
phendf <- dplyr::left_join(phenology_data, SPU_dat) %>%
    unique() %>%
    mutate(state = as.factor(Phenophase_Derived))

view(dfSummary(phendf))
# summary(phendf)

```

```{r stanindexing}
#Create indexes that stan will like
stanindexer <- function(df) {
    df$CloneID <- group_indices(df, Clone)
    df$OrchardID <- group_indices(df, Orchard)
    df$ProvenanceID <- group_indices(df, SPU_Name)
    df$SiteID <- group_indices(df, Site)
    df$YearID <- group_indices(df, Year)
    df$Tree <- group_indices(df,TreeID)
    return(df)
}

fdf <- filter(phendf, Sex == "FEMALE")
#fdf <- stanindexer(fdf)
mdf <- filter(phendf, Sex == "MALE")
#mdf <- stanindexer(mdf)

#test
nrow(fdf) + nrow(mdf) == nrow(phendf)

```

## Ristos or heatsum?

To answer this question, I can build models of each transition using ristos and using GDD.

Looking at the data, I don't really have a good sense of which is better.

```{r ristos and heatsum}
ggplot(phendf, aes(x=forcing_accum, fill=state)) +
    geom_histogram(alpha=0.5, binwidth = 10, position = "identity") +
    facet_grid(daily_forcing_type ~ Sex) +
    ggtitle("Histogram of forcing in each stage")

ggplot(phendf, aes(x=forcing_accum, color=state)) +
    stat_ecdf()+
    facet_grid(daily_forcing_type ~ Sex) +
    ggtitle("Cumulative forcing in each stage")




# ggplot(phendf, aes(x=forcing_accum, fill=state)) +
#     geom_density(alpha=0.5) +
#     facet_grid(daily_forcing_type ~ Sex) +
#     ggtitle("Forcing accummulations in each stage")
# 
# ggplot(mdf, aes(x=forcing_accum, fill=daily_forcing_type)) +
#     geom_density(alpha=0.5) +
#     facet_grid(SPU_Name ~ Site) + 
#     ggtitle("Phenophase Stage 2, Male")

```
Comparing the range of values each phase occurs over for ristos and GDD, ristos are generally higher. I believe there's more separation between phase 1 and 2 for GDD than ristos. For the transition from 2 to 3 I believe they are quite similar, though ristos may have more separation. Unclear what will happen with modelling.

```{r ristos and gdd}
ggplot(phendf, aes(x=state, y=forcing_accum, fill=daily_forcing_type)) +
    geom_violin()+
    facet_wrap(. ~ Sex) +
    ggtitle("Compare ristos and GDD required for each state")
```



```{r ristos v gdd}

#male
mgdd <- dplyr::filter(mdf, daily_forcing_type=="Heat")
mristo <- dplyr::filter(mdf, daily_forcing_type=="ristos")

fitmgdd <- bayespolr(factor(state) ~ forcing_accum, data = mgdd)
fitmristo <- bayespolr(state ~ forcing_accum, data=mristo)
AIC(fitmgdd, fitmristo)


#female
fgdd <- dplyr::filter(fdf, daily_forcing_type=="Heat")
fristo <- dplyr::filter(fdf, daily_forcing_type=="ristos")

fitfgdd <- bayespolr(factor(state) ~ forcing_accum, data = fgdd)
fitfristo <- bayespolr(state ~ forcing_accum, data=fristo)
AIC(fitfgdd, fitfristo)
```

It looks like growing degree days actually do a better job for the males and females. 

Is that true even after accounting for the largest effects? No! After accounting for effects of site and/or provenance and/or orchard, ristos are better. Clone and TreeID leave GDD doing better, but site and prov are the big effects.

```{r }
fitmgdd_sp <- bayespolr(state ~ forcing_accum + Site + SPU_Name, data=mgdd)
fitmristo_sp <- bayespolr(state ~ forcing_accum + Site + SPU_Name, data=mristo)
AIC(fitmgdd_sp, fitmristo_sp)

fitmgdd_sp <- bayespolr(state ~ forcing_accum + Site + SPU_Name, data=mgdd)
fitmristo_sp <- bayespolr(state ~ forcing_accum + Site + SPU_Name, data=mristo)
AIC(fitmgdd_sp, fitmristo_sp)
```

Sticking with ristos! Which is unfortunate because they are hard to explain.



## Varying slopes - transition state
Next question! Do slopes vary by transition state?

For this I will use logistic models for each transition in each sex.


```{r data prep do slopes vary}
#male
mdf <- filter(mdf, daily_forcing_type == "ristos")
mt1 <- filter(mdf, state %in% c(1,2)) %>% #male transition 1
    mutate(state = case_when(state==1 ~ 0,
                             state==2 ~ 1))

mt2 <- filter(mdf, state %in% c(2,3)) %>% #male transition 2
    mutate(state = case_when(state==2 ~ 0,
                             state==3 ~ 1))

#female

fdf <- dplyr::filter(fdf, daily_forcing_type == "ristos")
ft1 <- filter(fdf, state %in% c(1,2)) %>% #male transition 1
    mutate(state = case_when(state==1 ~ 0,
                             state==2 ~ 1))

ft2 <- filter(fdf, state %in% c(2,3)) %>% #male transition 2
    mutate(state = case_when(state==2 ~ 0,
                             state==3 ~ 1))
```

Model with logit glm.

```{r do slopes vary logits}
#male
fit_mt1 <- glm(state ~ forcing_accum, family = binomial(link = 'logit'), data = mt1)
fit_mt2 <- glm(state ~ forcing_accum, family = binomial(link = 'logit'), data = mt2)

fit_mt1$coefficients
fit_mt2$coefficients

get_slope_range <- function(fit, name) {
    low <- fit$coefficients[2] - se.coef(fit)[2]
    high <- fit$coefficients[2] + se.coef(fit)[2]
    return(c(model=name, lower=low, upper=high))
}

maleslopes <- rbind(get_slope_range(fit_mt1, "m1"),
                    get_slope_range(fit_mt2, "m2"))

#female
fit_ft1 <- glm(state ~ forcing_accum, family = binomial(link = 'logit'), data = ft1)
fit_ft2 <- glm(state ~ forcing_accum, family = binomial(link = 'logit'), data = ft2)

fit_ft1$coefficients
fit_ft2$coefficients

femaleslopes <- rbind(get_slope_range(fit_ft1, "f1"),
                      get_slope_range(fit_ft2, "f2"))

```

With the simplest version of the model, the slopes for the different transitions are definitely different.

`r maleslopes`
`r femaleslopes`

The first transition is slower than the second transition.

What if I include effects on the slope?

```{r }
#male
fit_met1 <- glm(state ~ forcing_accum*(Site+Clone+SPU_Name), family = binomial(link = 'logit'), data = mt1)
fit_met2 <- glm(state ~ forcing_accum*(Site+Clone+SPU_Name), family = binomial(link = 'logit'), data = mt2)

fit_met1$coefficients
fit_met2$coefficients

maleslopes <- rbind(get_slope_range(fit_met1, "m1"),
                    get_slope_range(fit_met2, "m2"))

AIC(fit_mt1, fit_met1)
AIC(fit_mt2, fit_met2)

#female
fit_fet1 <- glm(state ~ forcing_accum*(Clone), family = binomial(link = 'logit'), data = ft1)
fit_fet2 <- glm(state ~ forcing_accum*(Clone), family = binomial(link = 'logit'), data = ft2)

fit_fet1$coefficients
fit_fet2$coefficients

femaleslopes <- rbind(get_slope_range(fit_fet1, "f1"),
                      get_slope_range(fit_fet2, "f2"))

AIC(fit_ft1, fit_fet1)
AIC(fit_ft2, fit_fet2)
```

If I include a Site effect on the slope, the second transition is MUCH faster than the first transition.

Including Site, Clone, and Prov effects on slope brings them much closer together. True?

I don't know how to deal with an ordered logistic model where the transitions can have different slopes. Is that even an ordered logistic model anymore? I don't think so. Maybe I should just fit 4 models - each sex and each transition.

However, if I include Clone in the effect then 