#### posterior predictive checks
```{r depends}
library(tidyr)
library(dplyr)
library(stringr)
library(rstan)

```
```{r functions}

#from mcelreath's rethinking package
dordlogit <- function (x, phi, a, log = FALSE) 
{
    a <- c(as.numeric(a), Inf)
    p <- logistic(a[x] - phi)
    na <- c(-Inf, a)
    np <- logistic(na[x] - phi)
    p <- p - np
    if (log == TRUE) 
        p <- log(p)
    p
}
#from mcelreath's rethinking package
rorderedlogistic <- function (n, phi = 0, a) 
{
    a <- c(as.numeric(a), Inf)
    k <- 1:length(a)
    if (length(phi) == 1) {
        p <- dordlogit(k, a = a, phi = phi, log = FALSE)
        y <- sample(k, size = n, replace = TRUE, prob = p)
    }
    else {
        y <- rep(NA, n)
        if (n > length(phi)) {
            phi <- rep(phi, ceiling(n/length(phi)))
        }
        for (i in 1:n) {
            p <- dordlogit(k, a = a, phi = phi[i], log = FALSE)
            y[i] <- sample(k, size = 1, replace = TRUE, prob = p)
        }
    }
    y
}
```

```{r data and model}
#model
ffit.stan <- readRDS("female_slopes.rds")
mfit.stan <- readRDS("male_slopes.rds")

fmod <- as.data.frame(ffit.stan)
mmod <- as.data.frame(ffit.stan)


#data
phenology_data <- read.csv("data/stan_input/phenology_heatsum.csv",
                           stringsAsFactors = FALSE, header = TRUE) %>%
    filter(forcing_type=="ristos")

## provenance
SPU_dat <- read.csv("../research_phd/data/OrchardInfo/LodgepoleSPUs.csv",
                    header=TRUE, stringsAsFactors = FALSE) %>%
    dplyr::select(SPU_Name, Orchard)

# Data Processing ##################
# join provenance and phenology data

phendf <- phenology_data %>%
    na.omit()
phendf <- dplyr::left_join(phenology_data, SPU_dat) %>%
    unique()

# separate into male and female dataframes and turn factors into integers
fdf <- filter(phendf, Sex == "FEMALE")
fdf <- stanindexer(fdf)
mdf <- filter(phendf, Sex == "MALE")
mdf <- stanindexer(mdf)

# identify combinations of effects & predictor (sum_forcing) that actually occur
ufdf <- fdf %>%
    select(Site, SiteID, SPU_Name, ProvenanceID, Clone, CloneID, Year, YearID, sum_forcing) %>%
    distinct()
    
```

```{r get model df in long form}
#split by param

draws <- base::sample(1:nrow(fmod), 300)
fmod_sampled <- fmod[draws,]

singledimpars <- fmod_sampled %>% 
    mutate(draw=draws) %>%
    select(draw, beta, sigma_site, sigma_prov, sigma_clone, sigma_year, contains("kappa")) %>%
    rename(kappa1 = `kappa[1]`) %>%
    rename(kappa2 = `kappa[2]`)

tidypar <- function(stanframe, param, id) {
    #take a stan model dataframe and create a tidy dataframe for a given parameter. takes a dataframe, a string with the parameter name, and a string describing the param ID (e.g. par = "Site" and id="SiteID")
    par <- stanframe %>% select(contains(param)) %>%
        mutate(draw = draws) %>%
        gather(key = key, value = value, contains("b_")) %>%
        mutate(id = str_extract(key, "[0-9]{1,}")) 
    colnames(par) <- c("draw", "name", param, id)
   par[,4] <- as.integer(par[,4])
    return(par)
}

kappa <- select(fmod_sampled, contains("kappa"))
siteb <- tidypar(fmod_sampled, "b_site", "SiteID")
provb <- tidypar(fmod_sampled, "b_prov", "ProvenanceID")
cloneb <- tidypar(fmod_sampled, "b_clone", "CloneID")
yearb <- tidypar(fmod_sampled, "b_year", "YearID")

clonemerge <- left_join(ufdf, cloneb)
provmerge <- left_join(ufdf, provb) 
sitemerge <- left_join(ufdf, siteb)
yearmerge <- left_join(ufdf, yearb)

pardf <- data.frame(ufdf, 
                       draw = clonemerge$draw, 
                       b_clone = clonemerge$b_clone,
                       b_prov=provmerge$b_prov, 
                       b_site = sitemerge$b_site, 
                       b_year = yearmerge$b_year) %>% 
    left_join(singledimpars)

#pardf is a dataframe with N rows of parameters for each sum_forcing, site, provenance, clone, and year combination that appear in the data, where N = number of draws from the posterior distribution

```
# Now calculate state predictions 

```{r state predictions}

# try to generate states

pred_df <- pardf %>%
    mutate(betas = b_clone + b_prov + b_site + b_year + beta) %>%
    mutate(phi = betas * sum_forcing) 

state_pred <- c()
for (i in 1:nrow(pred_df)) {
    state_pred[i] <- rorderedlogistic(1, phi = pred_df$phi[i], a=c(pred_df$kappa1[i], pred_df$kappa2[i]))
}

pred_df <- data.frame(pred_df, state_pred=state_pred)

#merge in real data

real_state <- fdf %>% 
    dplyr::select(SiteID, ProvenanceID, Clone ID, YearID, sum_forcing, Phenophase_Derived))

expectations <- data.frame(sum_forcing=forcing_predictor, state=stateexp, source="predicted")

data <- filter(fdf, SiteID==1, ProvenanceID==1, YearID==10) %>%
    dplyr::select(sum_forcing, state=Phenophase_Derived) %>%
    mutate(source="data")

de <- rbind(expectations, data)

ggplot(de, aes(x=sum_forcing, color=as.factor(state), linetype=as.factor(source))) +
    stat_ecdf()

ggplot(de, aes(x=sum_forcing, color=as.factor(state), linetype=as.factor(source))) +
    geom_density()

mcmc_intervals(mpostdf, regex_pars = c("beta", "site", "prov")) + ggtitle("male")





