---
title: "Simulation comparisons"
author: "Susannah Tysor"
date: "April 19, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

This document describes simulations of phenology data, models used to analyze that data, results of those models, and comparisons of the model results.

## Simulations

Data is simulated using according to an underlying logistic model where the probability of transition from one phenological state to the next is a function of heatsum. The model has parameters `k` and `h`. `k` controls the slope of the transition curve. `h` controls the location of the transition curve - h is the inflection point. At `h`, half of the trees will have transitioned. 



```{r cars}
calc_probability <- function(x, k = steepness, h = midpoint, hi = threshold_ind, ki = steepness_ind) { # a logistic function parameterized to estimate steepness and transition points with individual effects
    1/(1 + exp(-(k + ki) * (x - (h + hi))))
}
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
