---
title: "A hierarchical Bayesian model of pollination phenology in Lodgepole pine"
author: "Susannah Tysor"
date: "March 19, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r depends}
library(rethinking)
library(dplyr)
library(bayesplot)

```

```{r stan setup}
util <- new.env()
source('~/Documents/classes/STANworkshop/material/1 - workflow/stan_utility.R', local=util)

options(mc.cores=parallel::detectCores())
rstan_options(auto_write=TRUE)
```
## Conceptual Analysis

Pollination phenology in lodgepole pine refers to the timing of two events - pollen shed and cone receptivity. I want to estimatethe beginning, end, and length of both events. 

As the developmental process for grown appears to be logistic, I will use a logistic model. There are two transitions I need to model with each 

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
